Evolution of ORM Systems: Formal Foundations

### 2013-01-09 doc/description.tex
has type of: \lhd

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented in an object-oriented language, which uses relational database as a data storage is context of this paper. The situation is illustrated in the Figure \ref{fig:appStructure}. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software a triple:
\begin{align*}
	software = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	application : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	database : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \in Applications, d \in Databases
\end{align*}
The software has to be in consistent state so it is beneficial for its user. The state of the software is defined to communicate the sate of the software to its user:
\begin{align*}
	state = Consistent\; s \: | \: Inconsistent \; s, \\
	s \in Softwares  \\ \\
	state : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases}
\end{align*}
\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), however the evolution has to be processed on the database level too. It means there is a set of application transformations called $AppOps$ and a set of database transformations called $DbOps$. Because there are needed different parameters for application and database evolution the set of transformation called $E$ which evolves the whole software and a mapping between $E$ and $AppOps$, respectively $DbOps$ is defined. We call this mappings $\Psi$ and $\Phi$:
\begin{align*}
\Psi : E \rightarrow AppOps \\
\Phi : E \rightarrow DbOps
\end{align*}
A subset of $E$ represents an evolution transformation of a software, which mets the first goal of our framework - a single source of data evolution for the whole software.

@@

\begin{align*}
 u \oplus t = \begin{cases}
 	\inference{u = Inconsistent \; s}{u} \\ \\
 	\inference{u = Consistent \; s}{state(sw(\Psi(u, t), \Phi(u,t))} 
\end{cases}\\ 
u \in States, t \in E, s \in Softwares
\end{align*}
\begin{align*}
 a \circ o = \begin{cases}
 	\inference{!\;consistent(a)}{a} \\ \\
 	\inference{consistent(a)}{a \circ o} 
\end{cases}\\ 
a \in Applications, o \in AppOps
\end{align*}
\begin{align*}
 d \bullet o = \begin{cases}
 	\inference{!\;consistent(d)}{d} \\ \\
 	\inference{consistent(d)}{d \circ o} 
\end{cases}\\ 
d \in Applications, o \in DbOps
\end{align*}

\section{Software Model}
The model of an ORM software consists of three parts - an application model, a database model and a ORM. The definitions of all parts assume a set $Label$ exist which contains all possible identifiers and each modeled entity is identified unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

@@

\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label, a = Application, cs = Class* 
\end{align*}

\paragraph{Class} represents a basic structural concept in the application model. It has a unique name, one or more properties and a class can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
% TODO association has to be only to self (all associated has to exists)
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{primitiveProperties} : Class \rightarrow Property* \\	primitiveProperties(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collectionProperties} : Class \rightarrow Property* \\	collectionProperties(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : reference(e) = l
  	}{ true } \\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*, a = Application
\end{align*}

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \in AppType, val \in String*, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as = Association
\end{align*}

@@

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{align*}
	\mathbf{Database} = <table*, row*> \\ \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	\mathbf{table} : Label \times Database \rightarrow Table  \\
	table(l, d) = t, t \in tables(d) \wedge name(t) = l \\ \\
	tabs = Table*, rs = Row*, l \in Label, d = Database
\end{align*}

\subsubsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\ \\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk = PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t = DbType, val = String,\\ cons = Constraints, c = Column, d = Database
\end{align*}

\paragraph{Foreign key}
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons = Constraint*, d = Database, fk = ForeignKey
\end{align*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i = Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
There functions $contains$ verify if the table contains data in context of concrete database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \in Database
\end{align*}
Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{align*}
	\mathbf{consistent} : Database \rightarrow Boolean \\
	consistent(d) = \begin{cases}
 		\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : \\ reference(p) = name(u)}{true}
 	\\ \\
 	false
 \end{cases}
\end{align*}

@@

\begin{algorithmic}[1]
	\Require $a \in Application$
	\State d = Database
	\ForAll{$c \in classes(a)$} 
		\State $\rho_c(c), d$
		\ForAll{$p \in properties(c)$} 
			\State $\rho_p(p, d)$
		\EndFor
	 \EndFor
	\ForAll{$c \in classes(a)$}
		\ForAll{$as \in associations(c)$} 
			\State $\rho_{a}(as, d)$
		\EndFor
	\EndFor
\end{algorithmic}

@@

\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		\inference{c = \varnothing}{\varnothing}\\\\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \\ \\
		\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))}
 \end{cases}\\
 c = Class
\end{align*}

@@

\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
 	\end{cases}
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}
\end{align*}

>>>>

\section{ORM Software}
\label{sec:evoIntro}
\begin{figure}
\centering
	\includegraphics[scale=0.3]{./images/system}
	\caption{The model of one generation of an ORM system consists of application persistent layer and relational database for storing instances.}
\label{fig:appStructure}
\end{figure}
A software implemented in an object-oriented language, which uses relational database as a data storage is context of this paper. The situation is illustrated in the Figure \ref{fig:appStructure}. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM (we will use $\rho$ as a symbol for ORM), therefore we define a software as a triple:
\begin{align*}
	\mathbf{software} = < Application, Database, \rho > \\
	sw : Applications \times Databases \times ORM \rightarrow Software \\ \\
	\mathbf{application} : Software \rightarrow Application \\
	application(sw(a, d, \rho)) = a \\ \\
	\mathbf{database} : Software \rightarrow Database \\
	database(sw(a, d, \rho)) = d \\
	a \lhd Application, d \lhd Database
\end{align*}
The software definition above introduce the first type used in this paper. The notation $a \lhd Application$ indicates the variable $a$ has the type $Application$.

The software has to be in consistent state so it is beneficial for its user. The state of the software is defined to communicate the sate of the software to its user:
\begin{align*}
 	\mathbf{state} = Consistent\; s \: | \: Inconsistent \; s, \\\\
	\mathbf{state} : Software \rightarrow State \\
	state(s) = \begin{cases}
 		\inference{consistent(application(s)) \wedge \\ consistent(database(s)) \\ \wedge \rho(application(s)) = database(s)}{Consistent \; s}\\ \\
 		\inference{otherwise}{Inconsistent \; s}
 	\end{cases} \\ \\
 	s \lhd Software  
\end{align*}
\subsection{Software Evolution}
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. We focus on situations when the evolution is initiated on the application level (as code refactoring), however the evolution has to be processed on the database level too. It means there is a set of application transformations called $AppOps$ and a set of database transformations called $DbOps$. Because there are needed different parameters for application and database evolution the set of transformation called $E$ which evolves the whole software and a mapping between $E$ and $AppOps$, respectively $DbOps$ is defined. We call this mappings $\Psi$ and $\Phi$:
\begin{align*}
\Psi : E \rightarrow AppOps \\
\Phi : E \rightarrow DbOps
\end{align*}
A subset of $E$ represents an evolution transformation of a software, which mets the first goal of our framework - a single source of data evolution for the whole software.

@@

\begin{align*}
 u \oplus t = \begin{cases}
 	\inference{u = Inconsistent \; s}{u} \\ \\
 	\inference{u = Consistent \; s}{state(sw(\Psi(u, t), \Phi(u,t))} 
\end{cases}\\ 
u \lhd State, t \in E, s \lhd Software
\end{align*}
\begin{align*}
 a \circ o = \begin{cases}
 	\inference{!\;consistent(a)}{a} \\ \\
 	\inference{consistent(a)}{a \circ o} 
\end{cases}\\ 
a \lhd Application, o \in AppOps
\end{align*}
\begin{align*}
 d \bullet o = \begin{cases}
 	\inference{!\;consistent(d)}{d} \\ \\
 	\inference{consistent(d)}{d \circ o} 
\end{cases}\\ 
d \lhd Application, o \in DbOps
\end{align*}

\section{Software Model}
The model of an ORM software consists of three parts - an application model, a database model and a ORM. The model of application and database creates a type system for both kinds ORM and evolutionary transformations.

The definitions of all parts assume a set $Label$ exist which contains all possible identifiers and each modeled entity is identified unambiguously identified in the model. Next there is a set $Values$ containing all possible values, which can be used in the model.

@@

\begin{align*}
  \mathbf{Application} = < Class* > \\ 
  \mathbf{app} : Class* \rightarrow  Application \\ \\
  \mathbf{classes} : Application \rightarrow Class* \\
  classes(app(cs)) = cs  \\\\
  \mathbf{class} : Label \times Application \rightarrow Class   \\ 
  class(l, a) = c, c \in classes(a) \wedge name(c) = l \\ \\
  l \in Label,a \lhd Applications, cs = Class* 
\end{align*}

\paragraph{Class} represents a basic structural concept in the application model. It has a unique name, one or more properties and a class can be associated to other classes in the application. 
\begin{align*}
	\mathbf{Class} = <label, Property*, Association*> \\
	\mathbf{cl} : Label \times Property* \times Association* \rightarrow Class \\\\
% TODO association has to be only to self (all associated has to exists)
	\mathbf{name} : Class \rightarrow Label \\
	name(cl(l, props, assocs)) = l \\ \\
	\mathbf{properties} : Class \rightarrow Property* \\
	properties(cl(l, props, assocs)) = props \\ \\
	\mathbf{primitiveProperties} : Class \rightarrow Property* \\	primitiveProperties(cl(l, props, assocs)) = props', \\ props' \subset props,  \forall p \in props' : cardinality(p) \leq 1 \\ \\
	\mathbf{collectionProperties} : Class \rightarrow Property* \\	collectionProperties(cl(l, props, assocs)) = props', \\ props' \subset props, \forall p \in props' : cardinality(p) > 1 \\ \\
	\mathbf{associations} : Class \rightarrow Association* \\	associations(cl(l, props, assocs)) = assocs \\ \\
	\mathbf{associated} :  Label \times Application \rightarrow Boolean \\
  	associated(l,a) = \begin{cases}
  		\inference{\exists c \in classes(a) \wedge \exists e \in associations(c) : reference(e) = l
  	}{ true } \\
 		false
 	\end{cases} \\ \\
	l \in Label, props = Property*, assocs = Association*,a \lhd Applications
\end{align*}

\paragraph{Property} represents a feature of  a class which is represented as a primitive type. The property can be mandatory, can have a default value and can represent a single value or a collection of values.
\begin{align*}
	\mathbf{Property} = <label, AppType, DefaultValue, Cardinality, Mandatory> \\ \\
	\mathbf{prop} : Label \times AppType \times String \times \mathbb{N^{*}} \times Boolean \rightarrow Property \\ \\
	\mathbf{name} : Property \rightarrow Label \\
	name(prop(l, t, val, n, m)) = l \\ \\
	\mathbf{type} : Property \times AppType \\
	type(prop(l, t, val, n, m)) = t \\ \\
	\mathbf{cardinality} : Property \rightarrow \mathbb{N^{*}} \\
	cardinality(prop(l, t, val, n, m)) = n \\ \\
	\mathbf{mandatory} : Property \rightarrow Boolean \\
	mandatory(prop(l, t, val, n, m)) = m  \\ \\
	\mathbf{owningClass} : Property \times Application \rightarrow Class  \\ 	
	owningClass(p, a) = \inference{\exists c \in classes(a) \wedge p \in properties(c)}{c}  \\ \\
	l \in Label, t \lhd AppType, val \in Values, n \in \mathbb{N^{*}}, m \in Boolean 
\end{align*}

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be the starting class of an association, referenced class is consider to be the ending class of an association. The cardinalities defines the multiplicity of the association ends.
\begin{align*}
	\mathbf{Association} = <label, classRef, startCardinality, endCardinality> \\ \\
	\mathbf{assoc} : Label \times Label \times \mathbb{N} \times \mathbb{N} \rightarrow Association\\
	\mathbf{name} : Association \rightarrow Label \\
	name(assoc(l, ref, n_1, n_2)) = l\\ \\
	\mathbf{reference} : Association \rightarrow Label \\
	reference(assoc(l, ref, n_1, n_2)) = ref\\ \\
	\mathbf{startCardinality} : Association \rightarrow \mathbb{N} \\
	startCardinality(assoc(l, ref, n_1, n_2)) = n_1\\ \\
	\mathbf{endCardinality} : Association \rightarrow \mathbb{N} \\
	endCardinality(assoc(l, ref, n_1, n_2)) = n_2 \\ \\
	\mathbf{owningClass} : Association \times Application \rightarrow Class  \\
	owningClass(as, a) = \inference{\exists c \in classes(a) \wedge as \in associations(c)}{c}  \\ \\	
	l, ref \in Label,  n_1, n_2 \in \mathbb{N}, as \lhd Association
\end{align*}

@@

\subsection{Database Model}
\label{sec:dbModel}
The relational database consists of database schema which defines the structure of the database and data, which in our ORM system represents instances. The database is:
\begin{align*}
	\mathbf{Database} = <table*, row*> \\ \\
	\mathbf{db} : Tables \times Rows \rightarrow Database \\ \\
	\mathbf{tables} : Database \rightarrow Tables \\
	tables(db(tabs, rs)) = tabs \\ \\
	\mathbf{rows} : Database \rightarrow Rows \\
	rows(db(tabs, rs)) = rs \\ \\
	\mathbf{table} : Label \times Database \rightarrow Table  \\
	table(l, d) = t, t \in tables(d) \wedge name(t) = l \\ \\
	tabs = Table*, rs = Row*, l \in Label,d \lhd Database
\end{align*}

\subsubsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
\begin{align*}
	\mathbf{Table} = <label, primaryKey, Column*, ForeignKey*>\\ \\
	\mathbf{tab} 	: Label \times PrimaryKeys \times Columns \times ForeignKeys \rightarrow Table \\ \\
	\mathbf{name} : Table \rightarrow Label  \\
	name(tab(l, pk, cols, fks)) = l \\ \\
	\mathbf{primaryKey} : Table \rightarrow PrimaryKeys  \\
	primaryKey(tab(l, pk, cols, fks)) = pk \\ \\
	\mathbf{columns} : Table \rightarrow Columns  \\
	columns(tab(l, pk, cols, fks)) = cols \\ \\
	\mathbf{foreignKeys} : Table \rightarrow ForeignKeys  \\
	foreignKeys(tab(l, pk, cols, fks)) = fks \\ \\
l \in Label, pk \lhd PrimaryKey, cols = Column*, fks = ForeignKey*
\end{align*}

\paragraph{Column} columns define possible data values and types which can be part of a table record.
\begin{align*}
	\mathbf{Column} = <label, DbType, DefaultValue, Constraint*> \\ \\
	\mathbf{col} : Label \times DbType \times String \times Constraints \rightarrow Column \\ \\
	\mathbf{name} : Column \rightarrow Label  \\
	name(col(l, t, val, cons)) = l  \\ \\
	\mathbf{type} : Column \rightarrow DbType  \\
	type(col(l, t, val, cons)) = t  \\ \\
	\mathbf{constraints} : Column \rightarrow Constraints  \\
	constraints(col(l, t, val, cons)) = cons  \\ \\
	\mathbf{owningTable} : Column \times Database \rightarrow Table  \\
	owningTable(c, d) = \inference{\exists tab \in tables(d) \wedge c \in columns(tab)}{tab} \\ \\
l \in Label, t \lhd DbType, val \in Values,\\ cons \lhd Constraints, c \lhd Column,d \lhd Database
\end{align*}

\paragraph{Foreign key}
\begin{align*}
	\mathbf{ForeignKey} = <label, tableRef, Constraint*> \\ \\
	\mathbf{fk} : Label \times Label \times Constraints \rightarrow ForeignKey \\ \\
	\mathbf{name} : ForeignKey \rightarrow Label \\
	name(fk(l, tRef, cons)) = lab  \\ \\
	\mathbf{reference} : ForeignKey \rightarrow Label  \\
	reference(fk(l, tRef, cons)) = tRef  \\ \\
	\mathbf{constaints} : ForeignKey \rightarrow Constraint*  \\
	constaints(fk(l, tRef, cons)) = cons  \\ \\
	\mathbf{owningTable} : ForeignKey \times Database \rightarrow Table  \\
	owningTable(fk, d) = \inference{\exists tab \in tables(d) \wedge fk \in foreignKeys(tab)}{tab}\\\\
	l, tRef \in Label, cons \lhd Constraint*,d \lhd Database, fk \lhd ForeignKey
\end{align*}

@@

\subsubsection{Model of Stored Data}
A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of reference to a concrete table and a set of value pairs, which represents concrete values of concrete columns.
\begin{align*}
	\mathbf{Row} = < refTable, pair* > \\
	\mathbf{row} : Label \times Pairs \rightarrow Row \\ \\
	\mathbf{refTable} : Row \rightarrow Label \\
	refTable(row(l, ps) = l \\ \\
	\mathbf{pairs} : Row \rightarrow Pair* \\
	pairs(row(l, ps)) = ps \\ \\
	\mathbf{pairByRef} : Label \times Row \rightarrow Pair \\
	pairByRef(l,i) = \begin{cases}
		\inference{\exists p \in pairs(i) : reference(p) = l}{p} \\ \\
			\varnothing
		\end{cases}\\
	 l \in Label, ps = Pair*, i \lhd Row
\end{align*}
The value stored in concrete column of the row is represented by pair, where the $refCol$ is a name of the column, a foreign key or its primary key and $value$ represents concrete value .
\begin{align*}
	\mathbf{Pair} = < refCol, value > \\
	[Label \times Value] \rightarrow Pair\\ \\
	\mathbf{reference} : Pair \rightarrow Label \\
	reference([l,val]) = l \\ \\
	\mathbf{value} : Pair \rightarrow Label \\ 
	value([l,val]) = val \\ \\
	l \in Label, val \in Value
\end{align*}
There functions $contains$ verify if the table contains data in context of concrete database.
\begin{align*}
	\mathbf{contains} : Table \times Database \rightarrow Boolean \\
	contains(t, d) = \begin{cases}
			\inference{\exists r \in data(d) : refTable(r) = name(t) }{true} \\ \\
		 	false
 		\end{cases} \\ \\
 l \in Label, ps = Pair*, t \in tables(d), d \lhd Database
\end{align*}
Each row in a database has to refer to a table in database schema and has a structure corresponding to the one defined by a table, otherwise the database is inconsistent. The consistency of database is then:
\begin{align*}
	\mathbf{consistent} : Database \rightarrow Boolean \\
	consistent(d) = \begin{cases}
 		\inference{\forall r \in data(d), \exists t \in tables(d) : reference(r) = name(t) \\ \wedge \forall p \in pairs(r) :\\ \exists u \in columns(t) \cup foreignKeys(t) : \\ reference(p) = name(u)}{true}
 	\\ \\
 	false
 \end{cases}
\end{align*}

@@

\begin{algorithmic}[1]
	\Require $a \in Application$
	\State $d \lhd Database$
	\ForAll{$c \in classes(a)$} 
		\State $\rho_c(c), d$
		\ForAll{$p \in properties(c)$} 
			\State $\rho_p(p, d)$
		\EndFor
	 \EndFor
	\ForAll{$c \in classes(a)$}
		\ForAll{$as \in associations(c)$} 
			\State $\rho_{a}(as, d)$
		\EndFor
	\EndFor
\end{algorithmic}

@@

\begin{align*}
	\mathbf{\rho_{c}} : Class*  \rightarrow Table* \\
	\rho_{c}(c) = \begin{cases}
		\inference{c = \varnothing}{\varnothing}\\\\
		\inference{|c| = 1}{table(name(c), primaryKey("Id"), \varnothing, \varnothing)}  \\ \\
		\inference{|c| > 1}{\rho_c(head(c)) \cup \rho_c(tail(c))}
 \end{cases}\\
 c \lhd Class
\end{align*}

@@

\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Column* \\ \\
	\rho_{sp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{sp}(head(p)) \cup \rho_{sp}(tail(p))}\\\\
 		\inference{cardinality(p) = 1 \wedge !\;mandatory(p)
 			\\ \wedge \exists t : t = table(name(owningClass(p)), d) 
		 }{\begin{gathered}
	  		column(name(p), \rho_t(type(p)), \varnothing )
		\end{gathered}} \\ \\
  		\inference{cardinality(p) = 1 \wedge mandatory(p)
			 \\ \wedge \exists t : t = table(name(owningClass(p)), d)
			\\ \wedge \not instanciated(owningClass(p))
		}{\begin{gathered}
 	 		column(name(p), \rho_t(type(p)), NOTNULL ) \\ 
		\end{gathered}}
 	\end{cases}
\end{align*}
\begin{align*}
	\mathbf{\rho_p} : Property* \rightarrow Table* \\ \\
	\rho_{cp}(p) = \begin{cases}
		\inference{p = \varnothing}{\varnothing}\\
		\inference{|p| > 1}{\rho_{cp}(head(p)) \cup \rho_{cp}(tail(p))}\\\\
     	\inference{cardinality(p) > 1 \wedge !\;mandatory(p) \\ 
     		\wedge 	\exists t : t = table(name(owningClass(p)), d)
     		}{\begin{gathered}
     			table(name(p), primaryKey("Id"), \\ column("value", 			\rho_t(type(p)), \varnothing), \\ foreignKey(name(p), 			name(owningClass(p)), NOTNULL) \\
    	\end{gathered}}\\ \\
     	\inference{ cardinality(p) > 1 \wedge mandatory(p) \\ 
     		\wedge \exists t : t = table(name(owningClass(p)), d) 
     		}{\begin{gathered}  
     			table(name(p), primaryKey("Id"), \\ 
     			column("value", \rho_t(type(p)), NOTNULL), \\ 			foreignKey(name(p), name(owningClass(p)), NOTNULL)
    	 \end{gathered}}
 	\end{cases}\\
 	p \lhd Property
\end{align*}
